Velocity Map Imaging Spectroscopy of C$_2$H$^-$ and C$_2$D$^-$: a benchmark study of vibronic coupling interactions

### 2022-04-12 Aug2021-CH2-.tex
fix figures

In this work the vibronic spectrum of the ethynyl radical C$_2$H is examined by employing High-Resolution Photoelectron Imaging (HR-PEI) in order to benchmark CFOUR~\cite{dev20}  \emph{ab-initio} vibronic-coupling calculations. C$_2$H is reported to be one of the most abundant molecules in the universe, and is the most thoroughly studied of the C$_{2n}$H species~\cite{wil91,hei99,for10}. Many different experimental techniques have been employed to try and resolve the complex vibronic spectrum, including electron spin resonance~\cite{coc64,jin85}, laser magnetic resonance~\cite{bro88,pfe96,sch98}, microwave and milli-meter wave spectroscopy~\cite{sas81,got83,end89,mul00}, infrared (matrix isolation~\cite{she87,jac87,for95} and Fourier Transform emission~\cite{ver88}) spectroscopy, laser induced fluorescence spectroscopy~\cite{hsu93,hsu95,chi99} and photoelectron spectroscopy~\cite{erv91,tay98,zho07}. C$_2$H has also received extensive theoretical attention to try and understand the experimental results, however the large number of possible vibronic levels means this has proven to be a challenge~\cite{per90,per91,per91b,per91c,per92,car00,tar03,tar04,for10,sta21}. In this work we demonstrate how construction of a quasidiabatic Hamiltonian allows for the strength of vibronic interactions between coupled surfaces near a conical intersection to be calculated, in order to simulate electronic and vibronic spectra. By employing anion HR-PEI both the $^2\Sigma^+$ and $^2\Pi$ surfaces are mapped out on an equal footing from the anion $\tilde{X}^1\Sigma^+$ state, allowing a direct comparison with \emph{ab-initio} modelling.

@@

\begin{table}
	\caption{Parameters for the quasidiabatic Hamiltonian Eq.~(\ref{eq:calc1}) determined using CFOUR and the quasidiabatic ansatz.}
	\label{tab:1}
	\begin{tabular}{c |c c c c| c c c c }
		 & C$_2$H & &  & & C$_2$D & & & \\
		 parameter & Anion & $\tilde{X}^2\Sigma^+$ & $\tilde{A}^2\Pi (a')$ & $\tilde{A}^2\Pi (a'')$ & Anion & $\tilde{X}^2\Sigma^+$ & $\tilde{A}^2\Pi (a')$ & $\tilde{A}^2\Pi (a'')$ \\
		 \hline\hline
		 Energy & 0 & 22431.21 &  26543.67 & 26543.67 & 0&	22431.21 & 26543.67	& 26543.67\\
		 f1 & 0	& -596.04 & 641.6 &	641.6 & 0& -596.04& 641.6 & 641.6\\
		 f2 & 0 & -1515.21 & 1257.07 & 1257.07 & 0 & -1401.53 & 1115.17 & 1115.17\\
		 f3 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
		 f4 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
         f11 & 3424.32 & 3487.03 & 3482.16 & 3482.16 & 2611.42 & 2636.57 & 2681.25 & 2681.25\\
         f12 & 0.00 & -28.32 & 31.46 & 31.46 & 0.00 & -48.38 & 55.40 & 55.40\\
         f22 & 1894.71 & 1830.83 & 2099.57 & 2042.21 & 1792.49 & 1747.62 & 1914.39 & 1914.39\\
         f33 & 607.09 & 87.57 & 1883.48 & 658.96 & 481.89 & 120.09 & 1503.88 & 473.55\\
         f44 & 607.09 & 87.57 & 658.96 & 1883.48 & 481.89 & 120.09 & 473.55 & 1503.88\\
         $\lambda$ & -1185.26 & & 770.77991 & -429.49481 & -1014.58 & & 620.701579 & -382.57 \\
         \end{tabular}
\end{table}

@@

\begin{figure}[th!]
	\includegraphics[width=0.6\textwidth]{figures/266-plot.pdf}
	\caption{Photoelectron spectrum of C$_2$H$^-$ at 266~nm, showing vibrationally excited $\tilde{A}$ state structure. The simulated spectra shown in green is calculated using the quasidiabatic ansatz, and convolved with Gaussians with a FWHM of 40~cm$^{-1}$ to match the lower VMI resolution at large electron kinetic energies.}
	\label{fig:4}
\end{figure}
To investigate the effects of coupling on the higher excited vibronic levels on the $\tilde{A} ^2\Pi$ surface, the calculations were also compared to the experimental photoelectron spectrum of C$_2$H$^-$ at 266~nm, presented in Figure~\ref{fig:4}. Gaussians were fit to the transitions from Figure~\ref{fig:3}(a) with a FWHM of 40~cm$^{-1}$ to match the lower resolution of the experimental data at high electron kinetic energies. Near 3,900~cm$^{-1}$ the electronic origin of the $\tilde{A} ^2\Pi$ state is split over 5 vibronic levels, as discussed above. A similar effect is also observed for the $3^1_0$ band, which becomes split over 3 vibronic levels labelled $l$, $n$, and $o$. The calculated spectrum correctly predicts the position and intensity of peaks $n$ and $o$, but appears to strongly underestimate the intensity of peak $l$. Between the $0^0_0$ and $3^1_0$ bands a collection of weaker peaks are also observed ($f-k$). These are assigned to highly excited HT coupled $\tilde{X} ^2\Sigma^+$ transitions ($j,k$), admixed $\tilde{X}-\tilde{A}$ transitions ($h,i$) and pure $\tilde{X} ^2\Sigma^+$ transitions ($f$). Consequently, peak $f$ will possess $\Sigma$ symmetry, and should have a different anisotropy to all of the other dominant transitions above 3,800~cm$^{-1}$. The position and intensity of these highly-excited coupled transitions are well described by the simulated spectrum.

@@

Photodetachment to the ground state of C$_2$H $(\tilde{X}\,^2\Sigma^+)$ involves ejecting an electron from an $s-$like $\sigma$ orbital (approximately $5\sigma_g$ in symmetry character), whereas detachment to the excited $\tilde{A}\,^2\Pi$ state occurs from a $p-$like $\pi$ orbital (approximately $1\pi_u$ in character). Therefore, the electron anisotropies may be described using the mixed $s-p$ model\cite{khu14},
\begin{equation}
\beta_{sp}(\epsilon) = \frac{2(1-\gamma_p)B_1\epsilon + \gamma_p(2A_1^2\epsilon^2-4A_1\epsilon\cos\delta_{2,0})}
{(1-\gamma_p)B_1\epsilon+\gamma_p(1+2A_1^2\epsilon^2)}
\label{eq:beta-sp}
\end{equation}
where $\epsilon$ is the electron kinetic energy and $\gamma_p$ is the fraction of $p$ character of the detachment orbital described as
\begin{equation}
|\psi\rangle = \sqrt{1-\gamma_p}|s\rangle + \sqrt{\gamma_p}|p\rangle.
\end{equation}
$A_1$ and $B_1$ in Eq.~(\ref{eq:beta-sp}) are the generalised Hanstorp~\cite{han89} coefficients describing the assumed Wigner-like~\cite{wig48} relative scalings of the radial transition dipole matrix elements for different allowed detachment channels. Specifically, $A_1\epsilon$ describes the energy-dependent ratio of the $p\rightarrow d$ and $p\rightarrow s$ transition amplitudes, while $B_1\epsilon$ corresponds to the $s\rightarrow p$ and $p\rightarrow s$ cross-section ratio~\cite{khu14}. It can be shown that under certain approximations $B_1/A_1 = 8/3$~\cite{san13}. Finally, $\delta_{2,0}$ in Eq.~(\ref{eq:beta-sp}) is the phase shift between the $s$ and $d$ partial waves, which in most cases of anion photodetachment is assumed to be small, corresponding to $\cos\delta_{2,0}\approx1$.

>>>>

In this work the vibronic spectrum of the ethynyl radical C$_2$H is examined by employing High-Resolution Photoelectron Imaging (HR-PEI) in order to benchmark CFOUR~\cite{dev20}  \emph{ab-initio} vibronic-coupling calculations. C$_2$H is reported to be one of the most abundant molecules in the universe, and is the most thoroughly studied of the C$_{2n}$H species~\cite{wil91,hei99,for10}. Many different experimental techniques have been employed to try and resolve the complex vibronic spectrum, including electron spin resonance~\cite{coc64,jin85}, laser magnetic resonance~\cite{bro88,pfe96,sch98}, microwave and milli-meter wave spectroscopy~\cite{sas81,got83,end89,mul00}, infrared (matrix isolation~\cite{she87,jac87,for95} and Fourier Transform emission~\cite{ver88}) spectroscopy, laser induced fluorescence spectroscopy~\cite{hsu93,hsu95,chi99} and photoelectron spectroscopy~\cite{erv91,tay98,zho07}. C$_2$H has also received extensive theoretical attention to try and understand the experimental results, however the large number of possible vibronic levels means this has proven to be a challenge~\cite{per90,per91,per91b,per91c,per92,car00,tar03,tar04,for10,sta21,gul21}. In this work we demonstrate how construction of a quasidiabatic Hamiltonian allows for the strength of vibronic interactions between coupled surfaces near a conical intersection to be calculated, in order to simulate electronic and vibronic spectra. By employing anion HR-PEI both the $^2\Sigma^+$ and $^2\Pi$ surfaces are mapped out on an equal footing from the anion $\tilde{X}^1\Sigma^+$ state, allowing a direct comparison with \emph{ab-initio} modelling.

@@

\begin{table}
	\caption{Parameters for the quasidiabatic Hamiltonian Eq.~(\ref{eq:calc1}) determined using CFOUR and the quasidiabatic ansatz.}
	\label{tab:1}
	\begin{tabular}{c |c c c c| c c c c }
		 & C$_2$H & &  & & C$_2$D & & & \\
		  & Anion & $\tilde{X}^2\Sigma^+$ & $\tilde{A}^2\Pi (a')$ & $\tilde{A}^2\Pi (a'')$ & Anion & $\tilde{X}^2\Sigma^+$ & $\tilde{A}^2\Pi (a')$ & $\tilde{A}^2\Pi (a'')$ \\
		 \hline\hline
		 $\Delta$E & 0 & 22431.21 &  26543.67 & 26543.67 & 0&	22431.21 & 26543.67	& 26543.67\\
		 f1 & 0	& -211.28 & 286.9 &	286.9 & 0& -596.04& 641.6 & 641.6\\
		 f2 & 0 & -1515.21 & 1257.07 & 1257.07 & 0 & -1401.53 & 1115.17 & 1115.17\\
		 f3 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
		 f4 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
         f11 & 3424.32 & 3487.03 & 3482.16 & 3482.16 & 2611.42 & 2636.57 & 2681.25 & 2681.25\\
         f12 & 0.00 & -28.32 & 31.46 & 31.46 & 0.00 & -48.38 & 55.40 & 55.40\\
         f22 & 1894.71 & 1830.83 & 2099.57 & 2042.21 & 1792.49 & 1747.62 & 1914.39 & 1914.39\\
         f33 & 607.09 & 87.57 & 1883.48 & 658.96 & 481.89 & 120.09 & 1503.88 & 473.55\\
         f44 & 607.09 & 87.57 & 658.96 & 1883.48 & 481.89 & 120.09 & 473.55 & 1503.88\\
         $\lambda$ &  & & -1185.26 &  &  & & -1014.58 &  \\
         \end{tabular}
\end{table}

@@

\begin{figure}[th!]
	\includegraphics[width=0.6\textwidth]{figures/266-plot.pdf}
	\caption{Photoelectron spectrum of C$_2$H$^-$ at 266~nm, showing vibrationally excited $\tilde{A}$ state structure. The simulated spectra shown in green is calculated using the quasidiabatic ansatz, and convolved with Gaussians with a kinetic energy dependent FWHM of $\Delta\text{E}/\text{E} = 0.04\%$ ($\Gamma=55-30$~cm$^{-1}$) to match the lower VMI resolution at large electron kinetic energies.}
	\label{fig:4}
\end{figure}
To investigate the effects of coupling on the higher excited vibronic levels on the $\tilde{A} ^2\Pi$ surface, the calculations were also compared to the experimental photoelectron spectrum of C$_2$H$^-$ at 266~nm, presented in Figure~\ref{fig:4}. Gaussians were fit to the transitions from Figure~\ref{fig:3}(a) with a FWHM of 40~cm$^{-1}$ to match the lower resolution of the experimental data at high electron kinetic energies. Near 3,900~cm$^{-1}$ the electronic origin of the $\tilde{A} ^2\Pi$ state is split over 5 vibronic levels, as discussed above. A similar effect is also observed for the $3^1_0$ band, which becomes split over 3 vibronic levels labelled $l$, $n$, and $o$. The calculated spectrum correctly predicts the position and intensity of peaks $n$ and $o$, but appears to strongly underestimate the intensity of peak $l$. Between the $0^0_0$ and $3^1_0$ bands a collection of weaker peaks are also observed ($f-k$). These are assigned to highly excited HT coupled $\tilde{X} ^2\Sigma^+$ transitions ($j,k$), admixed $\tilde{X}-\tilde{A}$ transitions ($h,i$) and pure $\tilde{X} ^2\Sigma^+$ transitions ($f$). Consequently, peak $f$ will possess $\Sigma$ symmetry, and should have a different anisotropy to all of the other dominant transitions above 3,800~cm$^{-1}$. The position and intensity of these highly-excited coupled transitions are well described by the simulated spectrum.

@@

Photodetachment to the ground state of C$_2$H $(\tilde{X}\,^2\Sigma^+)$ involves ejecting an electron from an $s-$like $\sigma$ orbital (approximately $5\sigma_g$ in symmetry character), whereas detachment to the excited $\tilde{A}\,^2\Pi$ state occurs from a $p-$like $\pi$ orbital (approximately $1\pi_u$ in character)~\cite{gul21}. Therefore, the electron anisotropies may be described using the mixed $s-p$ model\cite{khu14},
\begin{equation}
\beta_{sp}(\epsilon) = \frac{2(1-\gamma_p)B_1\epsilon + \gamma_p(2A_1^2\epsilon^2-4A_1\epsilon\cos\delta_{2,0})}
{(1-\gamma_p)B_1\epsilon+\gamma_p(1+2A_1^2\epsilon^2)}
\label{eq:beta-sp}
\end{equation}
where $\epsilon$ is the electron kinetic energy and $\gamma_p$ is the fraction of $p$ character of the detachment orbital described as
\begin{equation}
|\psi\rangle = \sqrt{1-\gamma_p}|s\rangle + \sqrt{\gamma_p}|p\rangle.
\end{equation}
$A_1$ and $B_1$ in Eq.~(\ref{eq:beta-sp}) are the generalised Hanstorp~\cite{han89} coefficients describing the assumed Wigner-like~\cite{wig48} relative scalings of the radial transition dipole matrix elements for different allowed detachment channels. Specifically, $A_1\epsilon$ describes the energy-dependent ratio of the $p\rightarrow d$ and $p\rightarrow s$ transition amplitudes, while $B_1\epsilon$ corresponds to the $s\rightarrow p$ and $p\rightarrow s$ cross-section ratio~\cite{khu14}. It can be shown that under certain approximations $B_1/A_1 = 8/3$~\cite{san13}. Finally, $\delta_{2,0}$ in Eq.~(\ref{eq:beta-sp}) is the phase shift between the $s$ and $d$ partial waves, which in most cases of anion photodetachment is assumed to be small, corresponding to $\cos\delta_{2,0}\approx1$.
